NWR-YD: Pre-registration

### 2019-10-14 manuscript.Rmd
prereg in progress

---
title: "Non-word repetition among the Tsimane': A small-scale study"
shorttitle        : "NWR among the Tsimane'"

author: 
  - name          : "Alejandrina Cristia"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "29, rue d'Ulm, 75005, Paris, France"
    email         : "alecristia@gmail.com"
  - name          : "Steven Levinson"
    affiliation   : "2"
  - name          : "Marisa Casillas"
    affiliation   : "2"

affiliation:
  - id            : "1"
    institution   : "LSCP, Département d’études cognitives, ENS, EHESS, CNRS, Université PSL"
  - id            : "2"
    institution   : "MPI"

@@

Little is known about language development in children growing up in Rossel Island, a community of primarily subsistence farmers who tend to reside in clise-knitted villages where child care is distributed across many individuals, and who typically speak Yélî Dnyé, a phonologically and lexically complex language.

@@

- typical structure of NWR tasks: N of items
- complexity varies widely across tasks, sometimes within tasks on purpose
- some authors modulate structural complexity (nb syll, syllable structure), others familiarity via phonotactic probability
- calculation of frequency in Yêli is challenging bc low resource language
- so we only manipulated structural complexity, not probability
- previous work seems to avoid difficult sounds
- but we felt this was important to represent the language, so we also varied this factor

- previous work shows even 2yo can do the task
- but there are striking changes with age
- inconsistent results RE effects of SES (including parental education) & child sex

@@

- What is the accuracy (whole word, phoneme based, distance) overall?
- How does this change as a function of item complexity (nb syll, syll complexity, sound complexity)?
- How frequent are errors that result in real words? Is that a function of item complexity?
- Is individual variation explainable by child age, sex, parental education?

@@

1. Children are more accurate for mono-syllables than longer items
2. Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds
3. Children's accuracy increases with child age
4. Non-monolingual Yélî children are less accurate than monolingual ones when tested on the dominant language (we did not test any non-dominant language)
5. Previous NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated
6. To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperfomr later-born children

@@

1. The length distribution in Yélî words is more balanced than that in English, and thus the performance decline for poly- versus mono-syllables may be less pronounced than that for English. **Check for work on European languages that may have looked into this**
2. The Yélî sound inventory is very large and compressed, with many similar sounds that are acoustic and articulatory neighbors. Therefore, this may constitute a pressure for children to have finer aduitory skills (and perhaps more precise articulations) than children speaking languages with a simpler inventory. **no work looking at consonants & vowels? no work looking at nasal vowels in particular?**
3. We had no reservations that children's accuracy increases with child age for the Yélî community as well.
4. Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from non-kin (or at least from people other than their mothers, who tend to be the non-native speakers). If so, the difference between monolinguals and not monolinguals may be smaller than that found in other work. That said, one recent study shozs that most child-directed input in the first 2 years does come from the mother, so in so far as this input has a crucial formational role, then there may still be a performance gap between these two groups.
5. In the Rossel community, formal education plays an extremely minor role in ensuring individual's success, is not a good index of relative socio-economic status, and furthermore there is only a narrow range of variation in maternal educational attainment. This may lead to no or only very small advantages for children whose mothers are more educated, provided that the causal chain between maternal education and child language is via SES more broadly. However, if education directly boosts maternal verbal skills and the incidence of verbal behavior (as suggested by CITE), then we should still see a difference along this factor.
6. One main causal path between birth order and language development is via parental input (CITE). Given our arguments above for how mothers may not be as important among Rossel people than in other places, then the performance gap between first borns and later borns may be smaller.

@@

- The sample size is too small to have enough power to detect all of the relationships of interest
- In fact, the sample size is too small to detect some of the known effects (see Table XX **add table with r/ds from previous work**)
- By virtue of this being a language-specific study, it is unclear that our method (including saliently the items we used) is comparable in precision to previous NWR studies
- Analyses were designed and pre-registered by the same person who collected the data, who coded accuracy (albeit with a non-native ear) on the first 11 participants, and who segmented all children's production. Therefore, a key condition of blinding is not met

@@

Many  NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, always meeting the condition that they do not mean anything in the target language. We kept the same variation in item length and the non-meaningfulness requirement, but we did not vary syllable complexity and clusters because these are vanishingly rare in the language. We also increased the number of items and individual child would be tested on, so that a child would get up to 23 items to repeat, and we created more items and distributed them across children, so as to increase the coverage, and be able to study more items.

@@

A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yelî inventory to produce CV monosyllables. As before, items were automatically excluded if they appeared in the dictionary. Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out. Since the phonology and phonetics of Yélî is still in the process of being described (CITE), there could have been undocumented constraints that rendered items illegal. Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence). These candidates were presented to one informant, for a final check that they did not mean anything. Together with the 2018 selection, they were recorded using a headset XXX and an olympus XXX from the written form presented together with the same item orally. The complete recorded list was finally presented to two more informants, who could repeat all the items and who confirmed there were no real words.

@@

Therefore, we sought additional information by asking a local informant to listen through each child production, paried with the auditory target the child had been provided with, and make a judgment of whether the item was correctly or incorrectly repeated. We additionally asked her to transcribe exactly what the child said, providing some examples of the types of errors children in general make (without making specific reference to Yélî sounds or the items in the elicitation sets). This "native coding" provides both accuracy and qualtiative desciptions of what the child said that are likely more reliable than those of the non-native coder.

@@

read.csv("NWR-demo.csv",header=T)->demo
colnames(demo)
read.table("log2.txt",header=T,sep="\t")->x
x$rep=gsub(".wav","",gsub(".*_","",x$out),fixed=T)
x[order(x$out),]->x
x$cor.online=1
for(i in 1:dim(x)[1]) if(x$rep[i]>1) x$cor.online[i-1]<-0

@@

@@

We are grateful to informants and individuals who participated in the study. AC acknowledges financial and institutional support from Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004 ACLEW, ANR-14-CE30-0003 MechELex, ANR-17-EURE-0017) and the J. S. McDonnell Foundation Understanding Human Cognition Scholar Award.

>>>>

---
title: "Non-word repetition in Yélî Dnye: A small-scale study"
shorttitle        : "NWR in Yélî Dnye'"

author: 
  - name          : "Alejandrina Cristia"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "29, rue d'Ulm, 75005, Paris, France"
    email         : "alecristia@gmail.com"
  - name          : "Marisa Casillas"
    affiliation   : "2"
    address       : "Wundtlaan 1, 6525 XD, Nijmegen, The Netherlands"
    email         : "marisa.casillas@mpi.nl"

affiliation:
  - id            : "1"
    institution   : "LSCP, Département d’études cognitives, ENS, EHESS, CNRS, Université PSL"
  - id            : "2"
    institution   : "Max Planck Institute for Psycholinguistics"

@@

Little is known about language development in children growing up in Rossel Island, a community of primarily subsistence farmers who tend to reside in close-knitted villages where child care is distributed across many individuals, and who typically speak Yélî Dnyé, a phonologically and lexically complex language.

@@

- typical structure of NWR tasks: N of items
- complexity varies widely across tasks, sometimes within tasks on purpose
- some authors modulate structural complexity (nb syll, syllable structure), others familiarity via phonotactic probability
- calculation of frequency in Yêli is challenging because it is a low-resource language 
- so we only manipulated structural complexity, not probability, in the design of the stimuli
- previous work seems to avoid difficult sounds
- but we felt this was important to represent the language, so we also varied this factor

- previous work shows even 2yo can do the task
- but there are striking changes with age
- inconsistent results RE effects of SES (including parental education) & child sex
- bilingual disadvantage (could be due to lower exposure to input; learning 2 phonologies; learning 2 lexica)

@@

- What is the accuracy (whole word, phoneme based, distance) overall?
- How does this change as a function of item complexity (nb syll, syll complexity, sound complexity)?
- How frequent are errors that result in real words? Is that a function of item complexity?
- Is individual variation explainable by child age, sex, parental education, birth order, monolingual status?

@@

1. Children are more accurate for mono-syllables than longer items
2. Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds. Therefore, we expect higher accuracy for items with common sounds than in those with rare sounds.
3. Children's accuracy increases with child age.
4. Non-monolingual Yélî children are less accurate than monolingual ones when tested on the society-dominant language (we did not test any non-dominant language) 
<!-- **(MC: or just because they're acquiring two phonologies and so show slightly delayed linguistic development, in parallel with vocab development stuff; we don't have data on their dominant language, but i guess we can assume it's YD for most kids---i'm sure there are exceptions!)** ac2mc I added "society-" because that's what I meant (we didn't get proportions of time in which they hear YD versus other languages); I didn't add anything about why this would come out like this or not here, but I'll add this -- although I think the evidence on that is mixed (ie no strong evidence that learning 2 phonologies sets you back)-->
5. Previous NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated.
6. To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperform later-born children

@@

1. The length distribution in Yélî words is more balanced than that in English, and thus the performance decline for poly- versus mono-syllables may be less pronounced than that for English. **Check for work on European languages that may have looked into this**
2. The Yélî sound inventory is very large and compressed, with many similar sounds that are acoustic and articulatory neighbors. Therefore, this may constitute a pressure for children to have finer auditory skills (and perhaps more precise articulations) than children speaking languages with a simpler inventory. As a result, differences between easier and harder items may be smaller in this work than in other research. **no work looking at consonants & vowels? no work looking at nasal vowels in particular?**
3. We had no reservations that children's accuracy increases with child age for the Yélî community as well.
4. Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from people not in their nuclear family 
<!-- **(MC: definitely almost all kin; 5k of them)** ac2mc I'm not sure this is how bio anthropologists think about kin; because it's about how you attribute resources to boost your genes' chances of being passed on relative to those of others in the same space. So my guess is that there is still a difference between nuclear family and not. Rewrote kin -> nuclear family -->
(or at least from people other than their mothers, who tend to be the non-native speakers). If so, the difference between monolinguals and not monolinguals may be smaller than that found in other work
<!-- **(MC: maybe it'd be more useful to try and estimate the number of non-Y moms who live in each hamlet where testing is taking place?)** ac2mc not sure it's worth the effort, in any case other female adults are almost certain to have very little impact on the target child... Perhaps I'm not saying something very clearly? -->
. That said, one recent study on the same population shows that most child-directed input in the first 2 years does come from the mother 
<!-- **(MC: i can get you actual estimates for this for YD; let's go with that instead!)** -->
<!-- ac2mc I think I meant your own paper! added "on the same population "-->
, so in so far as this input has a crucial formational role, then there may still be a performance gap between these two groups.
5. In the Rossel community, formal education plays an extremely minor role in ensuring individual's success, is not a good index of relative socio-economic status, and furthermore there is only a narrow range of variation in maternal educational attainment. This may lead to no or only very small advantages for children whose mothers are more educated, provided that the causal chain between maternal education and child language is via SES more broadly. However, if education directly boosts maternal verbal skills and the incidence of verbal behavior (as suggested by CITE), then we should still see a difference along this factor.
6. One main causal path between birth order and language development is via parental input (CITE). Given our arguments above for how mothers may not be as important among Rossel people than in other places, then the performance gap between first borns and later borns may be smaller.

@@

- The sample size is too small to have enough power to detect all of the relationships of interest
- In fact, the sample size is too small to detect some of the known effects (see Table XX **add table with r/ds from previous work**)
- By virtue of this being a language-specific study, it is unclear that our method (including saliently the items we used) is comparable in precision to previous NWR studies
- Analyses were designed and pre-registered by the same person who collected the data, who coded accuracy (albeit with a non-native ear **(MC: i guess this isn't true anymore given all the Anna data?)**) on the first 11 participants, and who segmented all children's production. Therefore, a key condition of blinding is not met

@@

Many  NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, always meeting the condition that they do not mean anything in the target language. We kept the same variation in item length and the non-meaningfulness requirement **(MC: before i forget, note that one of the stims, 'yi', is in fact a word, and a pretty frequent one at that (meaning 'tree'); i'm not sure how that happened, sorry!)**, but we did not vary syllable complexity and clusters because these are vanishingly rare in the language. We also increased the number of items and individual child would be tested on, so that a child would get up to 23 items to repeat, and we created more items and distributed them across children, so as to increase the coverage, and be able to study more items.

@@

A second list of candidate items was generated in 2019 by selecting complex consonants and systematically crossing them with all the vowels in the Yelî inventory to produce CV monosyllables. As before, items were automatically excluded if they appeared in the dictionary. Additionally, since hearing vowel length in monosyllables in isolation is challenging, any item that had a short/long real word neighbor was filtered out **MC: i had the impression that you relaxed this short/long constraint in the end, but maybe i dreamt that :)**. Since the phonology and phonetics of Yélî is still in the process of being described (CITE), there could have been undocumented constraints that rendered items illegal **MC: well any intervocalic d that isn't a retroflex flap would be phonotactically illegal, i think---do we have any of those? i thought we maybe did but i don't remember and i don't remember, if so, what ghaalyu pronounced in his recording**. Therefore, we made sure that the precise consonant-vowel sequence occurred in some real word in the dictionary (i.e., that there was a longer word included the monosyllable as a subsequence). These candidates were presented to one informant, for a final check that they did not mean anything. Together with the 2018 selection, they were recorded using a headset XXX and an olympus XXX from the written form presented together with the same item orally. The complete recorded list was finally presented to two more informants, who could repeat all the items and who confirmed there were no real words.

@@

Therefore, we sought additional information by asking a local informant to listen through each child production, paried with the auditory target the child had been provided with, and make a judgment of whether the item was correctly or incorrectly repeated. We additionally asked her to transcribe exactly what the child said, providing some examples of the types of errors children in general make (without making specific reference to Yélî sounds or the items in the elicitation sets). This "native coding" provides both accuracy and qualtiative desciptions of what the child said that are likely more reliable than those of the non-native coder. **MC: if we have anna's codes for everything, do we need your coding for anything?**

@@

read.csv("NWR-demo.csv",header=T)->demo
colnames(demo)
subset(demo,included=="yes" & Age <12 & !is.na(Age))->inc

summary(inc)

read.table("log2.txt",header=T,sep="\t")->x
x$rep=gsub(".wav","",gsub(".*_","",x$out),fixed=T)
x[order(x$out),]->x
x$cor.online=1
for(i in 1:dim(x)[1]) if(x$rep[i]>1) x$cor.online[i-1]<-0

@@

### Exploratory analyses: The role of phonotactic frequency

**(MC: but we can try and do a cursory analysis based on the corpora we have from Steve and my transcription of naturalistic interactions!)**

@@

We are grateful to informants and individuals who participated in the study. AC acknowledges financial and institutional support from Agence Nationale de la Recherche (ANR-17-CE28-0007 LangAge, ANR-16-DATA-0004 ACLEW, ANR-14-CE30-0003 MechELex, ANR-17-EURE-0017) and the J. S. McDonnell Foundation Understanding Human Cognition Scholar Award. MC blabla NWO Veni Innovational Scheme Grant (XXX-XX-XXX).

### 2019-10-14 prereg.Rmd
prereg in progress

@@

<!-- 1.1.	List specific, concise, and testable hypotheses. Please state if the hypotheses are directional or non-directional. If directional, state the direction. A predicted effect is also appropriate here. If a specific interaction or moderation is important to your research, you can list that as a separate hypothesis.  -->
<!-- 1.2.	Example: If taste affects preference, then mean preference indices will be higher with higher concentrations of sugar.  -->
<!-- Our hypothesis is that child-centered rearing (as a proxy for relative level of child-directed speech) is positively related to size of phoneme inventories: we predict that in languages in which a higher proportion of adult speakers provide with more responses of “child-centered rearing” responses, we find larger phoneme inventories than for languages with more responses of “parents have a life of their own”.  -->

@@

@@

@@

@@

@@

@@

@@

### 1.	Sample size
<!-- 1.1.	Describe the sample size of your study. How many units will be analyzed in the study? This could be the number of people, birds, classrooms, plots, interactions, or countries included. If the units are not individuals, then describe the size requirements for each unit. If you are using a clustered or multilevel design, how many units are you collecting at each level of the analysis? -->
<!-- 1.2.	Example: Our target sample size is 280 participants. We will attempt to recruit up to 320, assuming that not all will complete the total task.  -->
<!-- 1.3.	More information: For some studies, this will simply be the number of samples or the number of clusters. For others, this could be an expected range, minimum, or maximum number. -->

### 1.	Sample size rationale
<!-- 1.1.	This could include a power analysis or an arbitrary constraint such as time, money, or personnel. -->
<!-- 1.2.	Example: We used the software program G*Power to conduct a power analysis. Our goal was to obtain .95 power to detect a medium effect size of .25 at the standard .05 alpha error probability.  -->
<!-- More information: This gives you an opportunity to specifically state how the sample size will be determined. A wide range of possible answers is acceptable; remember that transparency is more important than principled justifications. If you state any reason for a sample size upfront, it is better than stating no reason and leaving the reader to “fill in the blanks.” Acceptable rationales include: a power analysis, an arbitrary number of subjects, or a number based on time or monetary constraints -->

### 1.	Stopping rule
<!-- 1.1.	If your data collection procedures do not give you full control over your exact sample size, specify how you will decide when to terminate your data collection.  -->
<!-- 1.2.	Example: We will post participant sign-up slots by week on the preceding Friday night, with 20 spots posted per week. We will post 20 new slots each week if, on that Friday night, we are below 320 participants.  -->
<!-- 1.3.	More information: You may specify a stopping rule based on p-values only in the specific case of sequential analyses with pre-specified checkpoints, alphas levels, and stopping rules. Unacceptable rationales include stopping based on p-values if checkpoints and stopping rules are not specified. If you have control over your sample size, then including a stopping rule is not necessary, though it must be clear in this question or a previous question how an exact sample size is attained. -->

>>>>

In this study, we use nonword repetition (NWR), a technique that has been widely adopted to study the development of perception-production across multiple languages, particularly in Europe. In NWR studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in the language children are learning. The child should immediately try to say it back without changing anything. Accuracy is thought to reflect long-term phonological knowledge (which allows the child to perceive the item accurately even though it is not a real word they have encountered before) as well as online working memory (to encode the item in the interval between hearing it and saying it back). In our implementation of NWR, we additionally made sure that some of the items contained typologically rare and/or  challenging sounds, to additionally capture children's auditory-articulatory accuracy.

Little is known about language development in children growing up in Rossel Island, a community of primarily subsistence farmers who tend to reside in close-knitted villages where child care is distributed across many individuals, and who typically speak Yélî Dnyé, a phonologically and lexically complex language.

@@

<!-- 1.1.	List specific, concise, and testable hypotheses. Please state if the hypotheses are directional or non-directional. If directional, state the direction. A predicted effect is also appropriate here. If a specific interaction or moderation is important to your research, you can list that as a separate hypothesis.  -->
<!-- 1.2.	Example: If taste affects preference, then mean preference indices will be higher with higher concentrations of sugar.  -->
Based on previous work, we made the following predictions:

1. **Length:** Children are more accurate for mono-syllables than longer items
2. **Complexity:** Similarly, we do not know of NWR research that manipulates the difficulty of the sounds that are included in the items, but word naming and other research suggests that children are more accurate when producing easy and/or typologically common sounds than difficult and/or typologically rare sounds
3. **Age:** Children's accuracy increases with child age
4. **Monolingual status:** Non-monolingual Yélî children are less accurate than monolingual ones when tested on the society-dominant language (we did not test any non-dominant language) 
<!-- **(MC: or just because they're acquiring two phonologies and so show slightly delayed linguistic development, in parallel with vocab development stuff; we don't have data on their dominant language, but i guess we can assume it's YD for most kids---i'm sure there are exceptions!)** ac2mc I added "society-" because that's what I meant (we didn't get proportions of time in which they hear YD versus other languages); I didn't add anything about why this would come out like this or not here, but I'll add this -- although I think the evidence on that is mixed (ie no strong evidence that learning 2 phonologies sets you back)-->
5. **Maternal education:** Previous NWR evidence on this is mixed, but general findings on language development suggest that children whose mothers are more educated are more accurate than children whose mothers are less educated
6. **Birth order:**  To our knowledge, there is no previous NWR work on this, but other research suggests that first-born children should outperform later-born children

Although these predictions were made based on previous work, we also thought reasonable that they would not obtain for the Rossel children in particular, as follows:

1. The length distribution in Yélî words is more balanced than that in English, and thus the performance decline for poly- versus mono-syllables may be less pronounced than that for English. 
2. The Yélî sound inventory is very large and compressed, with many similar sounds that are acoustic and articulatory neighbors. Therefore, this may constitute a pressure for children to have finer auitory skills (and perhaps more precise articulations) than children speaking languages with a simpler inventory. 
3. We had no reservations that children's accuracy increases with child age for the Yélî community as well.
4. Anecdotally Yélî children grow up in close-knitted communities and thus may receive significant portions of their language input from people not in their nuclear family 
(or at least from people other than their mothers, who tend to be the non-native speakers). If so, the difference between monolinguals and not monolinguals may be smaller than that found in other work. That said, one recent study on the same population shows that most child-directed input in the first 2 years does come from the mother, so in so far as this input has a crucial formational role, then there may still be a performance gap between these two groups.
5. In the Rossel community, formal education plays an extremely minor role in ensuring individual's success, is not a good index of relative socio-economic status, and furthermore there is only a narrow range of variation in maternal educational attainment. This may lead to no or only very small advantages for children whose mothers are more educated, provided that the causal chain between maternal education and child language is via SES more broadly. However, if education directly boosts maternal verbal skills and the incidence of verbal behavior, then we should still see a difference along this factor.
6. One main causal path between birth order and language development is via parental input. Given our arguments above for how mothers may not be as important among Rossel people than in other places, then the performance gap between first borns and later borns may be smaller.

@@

This is both an experiment and an observational study. Questions pertaining to stimulus properties (length & complexity) can be considered experimental; the others bear on observing children's behavior albeit in a very constrained situation (age, monolingual status, maternal education, birth order).

@@

Blinding was partial, as follows:

- Personnel who interacted directly with the participants did know whether an item was one or multiple syllables long, and sometimes of other features of the child (whether they were monolingual, their age), but was not aware of the finer-grained hypotheses.
- Participants were not aware of the specific hypotheses

@@

Regarding the experimental section (length & complexity), this is a within-participant repeated measures design. Regarding the observational section (age, monolingual status, maternal education, birth order), this is a cross-sectional design.

@@

Regarding the experimental section (length & complexity), we did not randomize length but we did randomize complexity. Since the test is very short overall, it is unlikely that there would be a tiredness effect leading to lower performance towards the end. An improvement due to practice may have occurred. Since there are several items of the same length in a row, we can test for practice effect in such sections. The observational section (age, monolingual status, maternal education, birth order) was not overtly randomized; children were tested based on availability.

@@

Registration prior to analysis of the data: As of the date of submission, the data exist and you have accessed it, though no analysis has been conducted related to the research plan (including calculation of summary statistics).

Please note that an analysis of a different aspect of the data were carried out: online coding by a non-native speaker, of 11 out of the 30+ participants. These only inspected accuracy as a function of age.

@@

The data were collected by AC, and the coding by a native speaker was overseen by MC.

@@

Hamlets were visited and the word about the study was spread, so that children could come to be tested in exchange of a packet of biscuits. We asked for children aged between 4 and 10 years, but sometimes older children participated. Some children were coerced by parents or others to participate, which was obvious in that they did not repeat even real words or the did not repeat the prompts. Sometimes a crowd gathered around the study and children were shy, talking too low for the recording to be reliable. Therefore, our inclusion and exclusion criteria are:

- They must have participated freely
- They need to have repeated at least the first three trial items in order to be included
- Children must be 11 years of age or younger

### Sample size
<!-- 1.1.	Describe the sample size of your study. How many units will be analyzed in the study? This could be the number of people, birds, classrooms, plots, interactions, or countries included. If the units are not individuals, then describe the size requirements for each unit. If you are using a clustered or multilevel design, how many units are you collecting at each level of the analysis? -->
<!-- 1.2.	Example: Our target sample size is 280 participants. We will attempt to recruit up to 320, assuming that not all will complete the total task.  -->
<!-- 1.3.	More information: For some studies, this will simply be the number of samples or the number of clusters. For others, this could be an expected range, minimum, or maximum number. -->

We tested all children who made themselves available during our visits to different hamplets, which were 55. Application of our inclusion criteria led to a sample size of 40.

### Sample size rationale
<!-- 1.1.	This could include a power analysis or an arbitrary constraint such as time, money, or personnel. -->
<!-- 1.2.	Example: We used the software program G*Power to conduct a power analysis. Our goal was to obtain .95 power to detect a medium effect size of .25 at the standard .05 alpha error probability.  -->
<!-- More information: This gives you an opportunity to specifically state how the sample size will be determined. A wide range of possible answers is acceptable; remember that transparency is more important than principled justifications. If you state any reason for a sample size upfront, it is better than stating no reason and leaving the reader to “fill in the blanks.” Acceptable rationales include: a power analysis, an arbitrary number of subjects, or a number based on time or monetary constraints -->

Our sample size was motivated entirely for logistic reasons. Given this constraint, we revisited the literature to better understand which of the effects of interest we may be able to find or not.

To begin with, we looked at previous work. NWR studies vary in terms of how many children are included. Farabolini (2018) carried out a study of the literature resulting in two summary tables. The first summary table reviews literature on vocabulary-NWR correlations among bilinguals, showing that in 6 studies, the number of participants ranged between 19 and 105, with a mean of 48.7 and a median of 35.5. The second table reviews 13 studies on correlations between language exposure and NWR among bilinguals, with a median sample size of 52, mean 68, range 39-156. Our study is clearly on the small side in this literature.

Second, we looked for the size of effects in our research questions:

```{r len}
# Vance 2005 - 5yo because midpoint of their distribution
m1syll=76.3 
sd1syll=8.4

m3syll=73.5 
sd3syll=12.0

```{r comp}
# Kapalkova high versus low wordlike -- 4yo midpoint
m1difficult=15.57 
sd1difficult=3.21

m3easy=17.57  
sd3easy=2.57

```{r age}
# Vance 2005 -- 5 vs 7 because we also have these
m5yov=77.08
sd5yov=5.92

m7yov=81.42
sd7yov=7.77

# Kapalkova  -- 4 vs 5 although we only have a few of these
m4yok=13.05 
sd4yok=3.02

m5yok=13.6
sd5yok=2.64

```{r monobi}
# Lotem & Chiat
bil1.m=.86
bil1.sd=.09

mon1.m=.85
mon1.sd=.09

bil2.m=.82
bil2.sd=.13

mon2.m=.8
mon2.sd=.14

1. **Length:**
2. **Complexity:** 
3. **Age:** 
4. **Monolingual status:** 
5. **Maternal education:** 
6. **Birth order:**

### Stopping rule
<!-- 1.1.	If your data collection procedures do not give you full control over your exact sample size, specify how you will decide when to terminate your data collection.  -->
<!-- 1.2.	Example: We will post participant sign-up slots by week on the preceding Friday night, with 20 spots posted per week. We will post 20 new slots each week if, on that Friday night, we are below 320 participants.  -->
<!-- 1.3.	More information: You may specify a stopping rule based on p-values only in the specific case of sequential analyses with pre-specified checkpoints, alphas levels, and stopping rules. Unacceptable rationales include stopping based on p-values if checkpoints and stopping rules are not specified. If you have control over your sample size, then including a stopping rule is not necessary, though it must be clear in this question or a previous question how an exact sample size is attained. -->

Not relevant, our sample size is fixed.
